---
title: "Analysis on MPG VS automatic or manual transmission"
author: "Rdataman"
date: "Sunday, July 27, 2014"
output: pdf_document
---
```{r setoptions,echo=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE,cache=FALSE,results='hide' )
```
## 1. Executive Summary
This report explore how the MPG (miles per gallon) parameter can be influenced by the transmission type of a car.  
Detailed question is :   

1. Is an automatic or manual transmission better for MPG?
2. Quantifying how different is the MPG between automatic and manual transmissions?

Boxplot, t-test and multivariable regression model are performed to answer the questions. The conclusion is that a manual car tends to be 1.809 miles/gallon higher than an automatic car if other parameters of the two car are same.

## 2. load the Data
```{r readdata}
## keep a copy of the original data
data(mtcars); str(mtcars); mtcars0 <- mtcars
mtcars$am <- as.factor(mtcars$am);levels(mtcars$am) <-c("Automatic", "Manual");
```
This dataframe contains 32 obs. of  11 numeric variables.   

## 3. Means test (ttest) to the questions
A boxplot shows that manual transmission cars do better in MPG than the automtic ones(check Appendix 1).**The answer can be validated with the t test.**
```{r ttest,results='hold' }
tt<-t.test(mtcars$mpg~mtcars$am); c(tt$p.value,tt$estimate)
```
Clearly, the mean mpg for automatic and manual transmission cars are 17.1 and 24.4 respectively. The p-value is much smaller than 0.05. We can  conclude that automatic cars has lower mpg compared with manual cars included in this dataframe. 

## 4. Regression model for the questions.
### Simple Linear Regression
```{r simplefit}
 summary(lm(mpg~am,mtcars))
```
The above code shows a simple model (mpg~am). It is poorly fitted (R-squared:0.36). The reason is that "mpg" are also highly related to other factors than am (check Appendix 2). In order to quantify the mpg difference better, a multivariable regression model should be made.   

### Multivariate regression model
A stepwise model selection will be helpful to determine the final multivariable.
```{r model}
## transfer some other cols into factors
mtcars$cyl <- as.factor(mtcars$cyl) ; mtcars$vs <- as.factor(mtcars$vs) 
mtcars$gear <- as.factor(mtcars$gear) ; mtcars$carb <- as.factor(mtcars$carb)
modfit<-lm(mpg~., data=mtcars)
library(MASS)
step <- stepAIC(modfit, direction="both");step$anova 
```
mpg ~ cyl + hp + wt + am is suggested. Make a new fitting.
```{r finalfit, results='hold'}
finalfit<-lm(mpg ~ cyl + hp + wt + am, data=mtcars);summary(finalfit)$coefficients
```
The final fitting seems ok judging by the value of R-squared(0.866) and the Residuals VS Fitted plots in appendix3. The Normal Q-Q plot indicates that the residuals are close to a normally distribution.    
Interpretion of the oefficient

1. cly increase from 4 to 6 to 8, mpg decrease by 3 and 2.2 mile/gallon, respectively
2. mpg will decrease slighly with hp increase.
3. wt increases per 1000lb, the mpg will decrease by 2.5 mile/gallon.
4. **The coefficient of "am" in the model is 1.809. A manual car tends to be 1.809 miles/gallon higher than an automatic car if other parameters of the two car are same.** 

## 5. Conclusion
A multivariable model was built. The 4 predictors are number of cylinders, weight, gross horsepower and transmission type. This model gives a good fitting and proves that a manual car runs 1.809 miles/gallon than an automatic one.  

\pagebreak

# Appendix

## Appendix 1: Boxplot of mpg VS am
```{r mpgVSam, fig.height=4, fig.width=4}
boxplot(mtcars$mpg ~ mtcars$am, data = mtcars, ylab="mpg",
        xlab="transmission",main="data from mtcars")
```

\pagebreak   

## Appendix 2 : The dependence of mpg on other parameters.
The relationships can be shown by a scatterplot of mtcars.
```{r pairs}
pairs(mtcars0, panel=panel.smooth, main="mtcars Data")
```
The cor() can also show how mpg is affected by all other parameters ("am" is not included) of a car.
```{r correlation, results='hold'}
# check cor between mpg and others (no "am")
cor(mtcars$mpg,mtcars0[,-1])
```

\pagebreak   

## Appendix 3 : Residuals plots of our model.
```{r residualsPlot, results='hold'}
par(mfrow=c(2,2)); plot(finalfit)
```
